Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-07-08 README.Rmd
revisions for 'accept with revision' decision, add renv

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# Careers in Ruins: Academic Archaeology Job Trends From 2013 – 2023

@@

This repository contains the data and code for our poster presented at the Society of American Archaeology Annual Meeting in New Orleans 17-21 April 2024:

@@

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

>>>>

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

@@

This repository contains the data and code for our paper:

> Ben Marwick, Anne Marie Poole, Ailin Zhang, Setareh Shafizadeh, and Jess Beck (2025). Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology. _American Antiquity_

An earlier version was presented as a poster at the Society of American Archaeology Annual Meeting in New Orleans 17-21 April 2024:

@@

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `renv::install()` to ensure you have the packages this analysis depends on 
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### 2025-07-08 analysis/paper/article.qmd
revisions for 'accept with revision' decision, add renv

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. Tenure-track positions are a type of academic job that typically involve a probationary period (e.g. 5–7 years) culminating in an evaluation of research, teaching, and service, that if successful, results in permanent employment, job security and academic freedom. This obsessive focus on career trajectories is warranted. As of 2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions [@the20222022]; today, 71% of faculty in the U.S. are non-tenure-track [@culver]. The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors [@beck], including decreasing federal support for higher education, concomitant increases in student debt [@gusterson2017], a pronounced shift in university investment from faculty to administration [@graeber2018, pp. 162–163], and the growing privatization and market orientation of scientific research [@mirowski2011].

@@

In American anthropology, the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education [@earle2014]. When combined with the institution of tenure, the ADEA exemption allowed faculty to stay in their posts for as long as they liked. The median age for faculty in the U.S. now ranks among the highest for all professions [@kaskie2016]. In tandem with the gradual de-investment in American higher education since 1980s [@mirowski2011] and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, for example, @Passalacqua_2018 found ratios ranging from 1.21 to 0.82 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, there is one tenure-track position in the U.S. for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. In engineering, Larson et al., have calculated that providing jobs for even 50% of PhD graduates would require the field to expand at an "improbable" rate of 14% per year [-@larson2014, pp.747]. As they emphasize, "... the system in many places is saturated, far beyond capacity to absorb new PhDs in academia at the rates they are being produced" [@larson2014, pp.749].

@@

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. Although some time has passed since the UC experiment, institutional attitudes to diversity are likely to remain in flux for the foreseeable future. The 2023 Supreme Court decision in Students for Fair Admissions v. Harvard, which effectively ended affirmative action at American colleges and universities, in tandem with the Trump administration’s 2025 targeting of DEI initiatives [@knox2025] foreshadow a shift in judicial and political currents that is reshaping institutional approaches to—and even definitions of—diversity. The 180° pivot over the past five years—from requiring carefully crafted diversity statements as a matter of course to abrupt and sweeping policy shifts rendering such documents obsolete—indicates that national and institutional commitments to diversity may be only as deep as the latest strategic plan.

@@

## Contextualising the Archaeology Academic Jobs Wiki as a data source

@@

# for those editors active for more than three months:
p_editor_life_distr_year <-
edits_for_each_year_tbl_span %>%
 filter(edit_span_years > 0.25) %>%
ggplot() +
  aes(as_date(first_edit_year),
      reorder(edit_name, first_edit_year)) +
  geom_segment(aes(x =   first_edit,
                   xend = last_edit),
               linewidth = 0.1,
               lineend = "butt") +
  theme_minimal(base_size = small_base_size) +
  labs(x = "", y = "Individual editors\n(usernames and IP addresses redacted)") +
  theme(axis.text.y=element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),)

# size of edits per year
p_edit_size_per_year <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_year) %>%
  summarise(edit_size_num = sum(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Sum of edits (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# typical edit size per year
p_median_edit_size_per_year <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_year) %>%
  # median, because sometimes there are 1-2 very large
  # page reformatting edits
  summarise(edit_size_num = median(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Median edit size (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 Bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year. G: Breakdown of self-reported Wiki user categories for each year."
#| fig-height: 5

@@

Longer-term editors who represent job seekers active for multiple seasons reflect one of the underdiscussed realities of the academic job market in archaeology—time spent on that market. Previous research on professional trajectories in academic archaeology suggests that early career researchers have a limited shelf life. Mackie and Rockwell, for example, identify a hiring plateau that occurs around seven years after candidates defend their PhDs [-@mackieMarketShareAccounting2023, p. 5]. However, too little time on the market also seems to have a negative effect on placement rates—"very few graduates obtain TT employment when they are all but dissertation (ABD) or even immediately after graduation" [@mackieMarketShareAccounting2023, p. 5]. These patterns are not restricted to the United States. In Brami et al.’s [-@bramiPrecariousFutureReflections2023] survey of 419 early career researchers in European archaeology, likelihood of holding a permanent position shifted relative to time since PhD. Of the respondents who were 1 year or less to 4 years post-PhD, only 3% (3/116) held permanent positions. These numbers increased slightly for participants who had been on the market for 5–7 years, with 14% (7/49) holding permanent positions; similarly, 18% (6/34) of respondents who were on the market for 8 years or more held permanent positions.

This trajectory in archaeology—where there appears to be a temporal "sweet spot" post-PhD for obtaining permanent academic positions—dovetails with a larger meta-analysis of the career trajectories of PhDs from the humanities and humanistic social sciences. @main2019 used longitudinal Graduate Education Survey (GES) data from the Andrew W. Mellon Foundation to analyse the career trajectories of over 5,000 PhDs from disciplines in the humanities and humanistic social sciences at three time points: six months post PhD, three years post PhD, and eight years post-PhD. The proportion of surveyed doctorates in tenure-track or tenured faculty positions increased with time since PhD—41% of the sample held a tenure-track position six months after defending, 64% held a tenure-track position three years after defending, and 68% held a tenure-track or tenured position eight years after defending. In keeping with broader patterns observed across the humanities and social sciences, the variable tenures of editors on the Archaeology Jobs Wiki may partially reflect the temporal patterns in hiring relative to job seekers' time since PhD.

@@

```{r}
#| label: fig-show-map-of-hiring-institution-and-cc
#| fig-cap: "A: Frequency of hiring institution by Carnegie classification. B: Inset shows map of the United States showing the count of tenure-track job ads posted by all insititutions in each state during 2013 -- 2023"

@@

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013–2023 include New York, Texas, Pennsylvania, and Florida. These top five states correspond to the top five most populous U.S. states, suggesting that rates of hiring are approximately proportional to population density. The top five states for job ads are also the five states with highest number of degrees awarded in Anthropology [@IPEDSNCES]. Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options.

@@

Our study of a decade of tenure-track job ads in archaeology in the U.S. reveals diverse dynamics in the demand for specialisations in topics, methods and geographic focus, and in the instructions to applicants. While these dynamics are familiar to scholars actively seeking faculty positions, we believe this is the first time they have been quantified at such a large scale within archaeology. Trends in job ads reflect broader shifts in intellectual and practical priorities concerning archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, may be affected by a variety of factors.

Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013–2014. The wiki’s emphasis on tenure-track jobs runs counter to well-established patterns in the academic job market, wherein short-term and contingent positions are fast outpacing "the last good job\[s\] in America" [@aronowitz2001]. Research across anthropological sub-fields has shown that the number of tenure-track positions advertised within the discipline has also declined. Analyzing the job ads posted in *Anthropology News* in 1999–2000 versus 2019–2020, for example, Gershon and Rachok [-@HelloTristesTropes2021] found that the number of tenure-track positions decreased by 25% between their two time slices. Similarly, in his examination of jobs ads posted on the Biological Anthropology Academic Job Wiki between 2010–2017, Passalacqua noted that approximately one quarter of the 474 ads posted were for adjunct or visiting staff [-@Passalacqua_2018, pp.773]. Though Passalacqua did not examine how the proportion of job types changed over time, he did underscore "a current diverging trend of decreasing academic job advertisements and increasing doctoral degrees in biological anthropology which could result in serious consequences for the discipline" [-@Passalacqua_2018, pp.773]. Such observations have a surprising longevity. Rogge raised the alarm nearly half a century ago, warning of the "maladaptive" consequences of the exponential rate of disciplinary growth and impending consequences for job seekers [@rogge1976, pp.839]. The dwindling number of permanent academic positions is not a trend unique to anthropology—as of 2019, the AAUP documented a 36% increase in contingent positions over the preceding 15 years (AAUP 2022), and as of 2022, contingent positions make up more than half of faculty positions in the United States [@colby2022annual] .

@@

It is no secret that tenure-track jobs have traditionally been considered the "gold standard" of career outcomes for PhD students in the social sciences and humanities. As Kelsky outlines in *The Professor Is In*, her influential and best-selling guide for PhD students pursuing faculty careers, "...most ranking graduate programs sill consider any PhD who doesn't land a tenure-track job a failure or an aberration... Graduate students absorb this value system and judge themselves harshly" [-@kelsky2015, pp. 11]. These rubrics of success have been common across the humanities and social sciences for over half a century. Evaluating the demise of the National Endowment for Humanities Funded "Program to ready PhDs for Careers in Business" (CIB), which aimed to facilitate humanities PhDs' transition to the corporate world in the late 1970s, Franczak points to the dissonance between participants' incentives and outcomes:

@@

Our data suggest that archaeology faculty at U.S. universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. Due to the 1964 Civil Rights Act which prohibits hiring based solely on race or ethnicity, however, it is illegal for U.S. universities use race or ethnicity as a primary factor in hiring. As a result, universities appear to have tailored the content of their job ads to focus on topics where they expected Black and Indigenous researchers to be most numerous, in an effort to ensure that these researchers would be well-represented in the pool of applicants.

@@

One bright spot for applicants is the decline in requests for names of recommenders in the initial application in recent years. This may be a response to recent criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all. Showing sensitivity to this burden, in 2020 the American Anthropological Association issued guidance to academic departments that letters of recommendation should not be requested in the initial application, but should only be required from short-listed candidates[@jobboar2020; @Youngling2020].

@@

To echo Paul Farmer, "all this is both interesting and horrible" [-@farmer2004, pp.316]. Contextualising our job ad findings in this sociopolitical tumult provokes a suite of new questions about academic futures. Many archaeologists now have urgent concerns about what topics are likely to be "safe" and lead to secure ongoing employment, considering the current political and funding landscape. Will states with stronger cultural and environmental regulations offer more stability in employment, and if so, which states are likely to be the safest? These are legitimate and crucial questions, but predicting the future goes beyond the scope of our examination of the past decade of hiring dynamics in academic archaeology. Extrapolating future trends is inherently uncertain, especially in a mercurial political and economic landscape that is experiencing tectonic shifts in bureaucracy, organization, and funding on a near-weekly basis. Our study does, however, point to at least four findings that have implications for the next ten years of hiring, despite the turbulent political climate of 2024–2025.

>>>>

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. Tenure-track positions are a type of academic job that typically involve a probationary period (e.g. five to seven years) culminating in an evaluation of research, teaching, and service, that if successful, results in permanent employment, job security and academic freedom. This obsessive focus on career trajectories is warranted. As of 2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions [@the20222022]; today, 71% of faculty in the U.S. are non-tenure-track [@culver]. The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors [@beck], including decreasing federal support for higher education, concomitant increases in student debt [@gusterson2017], a pronounced shift in university investment from faculty to administration [@graeber2018, pp. 162–163], and the growing privatization and market orientation of scientific research [@mirowski2011].

@@

In American anthropology, the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education [@earle2014]. When combined with the institution of tenure, the ADEA exemption allowed faculty to stay in their posts for as long as they liked. The median age for faculty in the U.S. now ranks among the highest for all professions [@kaskie2016]. In tandem with the gradual de-investment in American higher education since 1980s [@mirowski2011] and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, for example, @Passalacqua_2018 found ratios ranging from 1.21 to 0.82 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, there is one tenure-track position in the U.S. for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. In engineering, Larson et al., have calculated that providing jobs for even 50% of PhD graduates would require the field to expand at an "improbable" rate of 14% per year [-@larson2014, pp.747]. As they emphasize, "... the system in many places is saturated, far beyond capacity to absorb new PhDs in academia at the rates they are being produced" [@larson2014, pp.749].

@@

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. Although some time has passed since the UC experiment, institutional attitudes to diversity are likely to remain in flux for the foreseeable future. The 2023 Supreme Court decision in Students for Fair Admissions v. Harvard, which effectively ended affirmative action at American colleges and universities, in tandem with the Trump administration’s 2025 targeting of DEI initiatives [@knox2025] foreshadow a shift in judicial and political currents that is reshaping institutional approaches to—and even definitions of—diversity. The 180 degree pivot over the past five years—from requiring carefully crafted diversity statements as a matter of course to abrupt and sweeping policy shifts rendering such documents obsolete—indicates that national and institutional commitments to diversity may be only as deep as the latest strategic plan.

@@

## Contextualizing the Archaeology Academic Jobs Wiki as a data source

@@

# for those editors active for more than three months:
p_editor_life_distr_year <-
edits_for_each_year_tbl_span %>%
 filter(edit_span_years > 0.25) %>%
  mutate(first_edit_year = as_date(first_edit_year)) %>% 
ggplot() +
  aes(first_edit_year,
      reorder(edit_name, first_edit_year)) +
  geom_segment(aes(x =   first_edit,
                   xend = last_edit),
               linewidth = 0.1,
               lineend = "butt") +
  theme_minimal(base_size = small_base_size) +
  labs(x = "", y = "Individual editors\n(usernames and IP addresses redacted)") +
  theme(axis.text.y=element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())  +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y")

# size of edits per year
p_edit_size_per_year <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_year) %>%
  summarise(edit_size_num = sum(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Sum of edits (bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# typical edit size per year
p_median_edit_size_per_year <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_year) %>%
  # median, because sometimes there are 1-2 very large
  # page reformatting edits
  summarise(edit_size_num = median(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Median edit size (bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year. G: Breakdown of self-reported Wiki user categories for each year."
#| fig-height: 4

@@

Longer-term editors who represent job seekers active for multiple seasons reflect one of the underdiscussed realities of the academic job market in archaeology—time spent on that market. Previous research on professional trajectories in academic archaeology suggests that early career researchers have a limited shelf life. Mackie and Rockwell, for example, identify a hiring plateau that occurs around seven years after candidates defend their PhDs [-@mackieMarketShareAccounting2023, p. 5]. However, too little time on the market also seems to have a negative effect on placement rates—"very few graduates obtain TT employment when they are all but dissertation (ABD) or even immediately after graduation" [@mackieMarketShareAccounting2023, p. 5]. These patterns are not restricted to the United States. In Brami et al.’s [-@bramiPrecariousFutureReflections2023] survey of 419 early career researchers in European archaeology, likelihood of holding a permanent position shifted relative to time since PhD. Of the respondents who were one year or less than four years post-PhD, only 3% (3/116) held permanent positions. These numbers increased slightly for participants who had been on the market for five to seven years, with 14% (7/49) holding permanent positions; similarly, 18% (6/34) of respondents who were on the market for eight years or more held permanent positions.

This trajectory in archaeology—where there appears to be a temporal "sweet spot" post-PhD for obtaining permanent academic positions—dovetails with a larger meta-analysis of the career trajectories of PhDs from the humanities and humanistic social sciences. @main2019 used longitudinal Graduate Education Survey (GES) data from the Andrew W. Mellon Foundation to analyse the career trajectories of over 5,000 PhDs from disciplines in the humanities and humanistic social sciences at three time points: six months post-PhD, three years post-PhD, and eight years post-PhD. The proportion of surveyed doctorates in tenure-track or tenured faculty positions increased with time since PhD—41% of the sample held a tenure-track position six months after defending, 64% held a tenure-track position three years after defending, and 68% held a tenure-track or tenured position eight years after defending. In keeping with broader patterns observed across the humanities and social sciences, the variable tenures of editors on the Archaeology Jobs Wiki may partially reflect the temporal patterns in hiring relative to job seekers' time since PhD.

@@

```{r}
#| label: fig-show-map-of-hiring-institution-and-cc
#| fig-cap: "A: Frequency of hiring institution by Carnegie classification. B: Inset shows map of the United States showing the count of tenure-track job ads posted by all institutions in each state during 2013 -- 2023"

@@

The geographic distribution of hiring institutions is shown in @fig-show-map-of-hiring-institution-and-cc Panel B. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013–2023 include New York, Texas, Pennsylvania, and Florida. These top five states correspond to the top five most populous U.S. states, suggesting that rates of hiring are approximately proportional to population density. The top five states for job ads are also the five states with highest number of degrees awarded in Anthropology [@IPEDSNCES]. Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options.

@@

Our study of a decade of tenure-track job ads in archaeology in the U.S. reveals diverse dynamics in the demand for specializations in topics, methods and geographic focus, and in the instructions to applicants. While these dynamics are familiar to scholars actively seeking faculty positions, we believe this is the first time they have been quantified at such a large scale within archaeology. Trends in job ads reflect broader shifts in intellectual and practical priorities concerning archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, may be affected by a variety of factors.

Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013–2014. The wiki’s emphasis on tenure-track jobs runs counter to well-established patterns in the academic job market, wherein short-term and contingent positions are fast outpacing "the last good job\[s\] in America" [@aronowitz2001]. Research across anthropological sub-fields has shown that the number of tenure-track positions advertised within the discipline has also declined. Analyzing the job ads posted in *Anthropology News* in 1999–2000 versus 2019–2020, for example, Gershon and Rachok [-@HelloTristesTropes2021] found that the number of tenure-track positions decreased by 25% between their two time slices. Similarly, in his examination of jobs ads posted on the Biological Anthropology Academic Jobs Wiki between 2010–2017, Passalacqua noted that approximately one quarter of the 474 ads posted were for adjunct or visiting staff [-@Passalacqua_2018, pp.773]. Though Passalacqua did not examine how the proportion of job types changed over time, he did underscore "a current diverging trend of decreasing academic job advertisements and increasing doctoral degrees in biological anthropology which could result in serious consequences for the discipline" [-@Passalacqua_2018, pp.773]. Such observations have a surprising longevity. Rogge raised the alarm nearly half a century ago, warning of the "maladaptive" consequences of the exponential rate of disciplinary growth and impending consequences for job seekers [@rogge1976, pp.839]. The dwindling number of permanent academic positions is not a trend unique to anthropology—as of 2019, the AAUP documented a 36% increase in contingent positions over the preceding 15 years (AAUP 2022), and as of 2022, contingent positions make up more than half of faculty positions in the United States [@colby2022annual] .

@@

It is no secret that tenure-track jobs have traditionally been considered the "gold standard" of career outcomes for PhD students in the social sciences and humanities. As Kelsky outlines in *The Professor Is In*, her influential and best-selling guide for PhD students pursuing faculty careers, "...most ranking graduate programs still consider any PhD who doesn't land a tenure-track job a failure or an aberration... Graduate students absorb this value system and judge themselves harshly" [-@kelsky2015, pp. 11]. These rubrics of success have been common across the humanities and social sciences for over half a century. Evaluating the demise of the National Endowment for Humanities Funded "Program to ready PhDs for Careers in Business" (CIB), which aimed to facilitate humanities PhDs' transition to the corporate world in the late 1970s, Franczak points to the dissonance between participants' incentives and outcomes:

@@

Our data suggest that archaeology faculty at U.S. universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. Due to the Civil Rights Act of 1964 which prohibits hiring based solely on race or ethnicity, however, it is illegal for U.S. universities use race or ethnicity as a primary factor in hiring. As a result, universities appear to have tailored the content of their job ads to focus on topics where they expected Black and Indigenous researchers to be most numerous, in an effort to ensure that these researchers would be well-represented in the pool of applicants.

@@

One bright spot for applicants is the decline in requests for names of recommenders in the initial application in recent years. This may be a response to recent criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all. Showing sensitivity to this burden, in 2020 the American Anthropological Association issued guidance to academic departments that letters of recommendation should not be requested in the initial application, but should only be required from short-listed candidates [@jobboar2020; @Youngling2020].

@@

To echo Paul Farmer, "all this is both interesting and horrible" [-@farmer2004, pp.316]. Contextualizing our job ad findings in this sociopolitical tumult provokes a suite of new questions about academic futures. Many archaeologists now have urgent concerns about what topics are likely to be "safe" and lead to secure ongoing employment, considering the current political and funding landscape. Will states with stronger cultural and environmental regulations offer more stability in employment, and if so, which states are likely to be the safest? These are legitimate and crucial questions, but predicting the future goes beyond the scope of our examination of the past decade of hiring dynamics in academic archaeology. Extrapolating future trends is inherently uncertain, especially in a mercurial political and economic landscape that is experiencing tectonic shifts in bureaucracy, organization, and funding on a near-weekly basis. Our study does, however, point to at least four findings that have implications for the next ten years of hiring, despite the turbulent political climate of 2024–2025.
